### 2015-06-06 chapters/pitfalls.tex
minor updates to chapter 2

The other problem with visual display is related to the so-called \textsc{font
rendering}. Font rendering refers to the process of the actual positioning of
Unicode characters on a page of written text. This positioning is actually a
highly complex problem, and many things can go wrong in the process. Well-known
rendering problems, like proportional glyph size or ligatures are reasonably
well understood. However, even within the Latin script, the positioning of
diacritics relative to a base character is still a widespread problem.
Especially when more than one diacritic is supposed to be placed above (or
below) each other, this often leads to unexpected effects in many modern
software applications. The problems arising in Arabic and in many southeast
Asian scripts (like Devanagari or Burmese) are even much more complex. There are
basically three different approaches to font rendering. The most widespread is
Adobe's and Microsoft's \emph{OpenType }system. This approach makes it
relatively easy for font developers, as the font itself does not include all
details about the precise placement of individual characters. For those details,
additional script-descriptions are necessary. All of those systems can lead to
unexpected behavior.\footnote{For more details about OpenType, see
\url{http://www.adobe.com/products/type/opentype.html} and
\url{http://www.microsoft.com/typography/otspec/}. Additional systems for
complex text layout are, among others, Microsoft's DirectWrite
\url{https://msdn.microsoft.com/library/dd368038.aspx} and the open-source
project HarfBuzz \url{http://www.freedesktop.org/wiki/Software/HarfBuzz/}.}
Alternative systems are Apple's \emph{Apple Advanced Typography} (AAT) and the
open-source \emph{Graphite} system from the Summer Institute of Linguistics
(SIL).\footnote{More information about AAT can be found at
\url{https://developer.apple.com/fonts/}. SIL's Graphite is described in detail
at
\url{http://scripts.sil.org/cms/scripts/page.php?site\_id=projects\&item\_id=graphite\_home}.}
In both of these systems, a larger burden is placed on the description inside
the font.

@@

\begin{itemize}
	\item <{\fontspec{EversonMono}A}> \textsc{latin capital letter a}, at \uni{0041} 
	\item <{\fontspec{EversonMono}А}> \textsc{cyrillic capital letter a}, at \uni{0410} 
	\item <{\fontspec{EversonMono}Α}> \textsc{greek capital letter alpha}, at \uni{0391} 
	\item <{\fontspec{EversonMono}Ꭺ}> \textsc{cherokee letter go}, at \uni{13AA} 
	\item <{\fontspec{EversonMono}ᗅ}> \textsc{canadian syllabics carrier gho}, at \uni{15C5} 
	\item <{\fontspec{EversonMono}ᴀ}> \textsc{latin small letter capital a}, at \uni{1D00} 
	\item <{\fontspec{EversonMono}ꓮ}> \textsc{lisu letter a}, at \uni{A4EE} 
%	\item <{\fontspec{EversonMono}Ａ}> \textsc{fullwidth latin capital letter a}, at \uni{FF21} 
	\item <{\fontspec{EversonMono}𐊠}> \textsc{carian letter a}, at \uni{102A0} 
	\item <{\fontspec{EversonMono}𐌀}> \textsc{old italic letter a}, at \uni{10300} 
	\item <{\fontspec{EversonMono}𝖠}> \textsc{mathematical sans-serif capital a}, \uni{1D5A0} 
	\item <{\fontspec{EversonMono}𝙰}> \textsc{mathematical monospace capital a}, at \uni{1D670} 
\end{itemize}

@@

Consider for example the characters and following Unicode code points:
\begin{itemize}
	\def\labelenumi{\arabic{enumi}.} 
	\item <Å> \textsc{latin capital letter a with ring above} \uni{00C5} 
	\item <Å> \textsc{angstrom sign} \uni{212B}
	\item <Å> \textsc{latin capital letter a} \uni{0041}
	+ \textsc{combining ring above} \uni{030A}
\end{itemize}

@@

There are two combinations that deserve specific care because it is actually 
possible to produce identical characters in different ways, without them being 
canonical equivalent. First, there are the
combining characters designated as ``overlay'' in the Unicode Standard, like    
<\ \ {\large  ̴}> \textsc{combining tilde overlay} at \uni{0334} or 
<\ \ {\large  ̵}> \textsc{combining short stroke overlay} at \uni{0335}. 
There are many characters that look like they are 
precomposed with such an overlay, for example 
<{\small \fontspec{CharisSIL}ɫ ᵬ ᵭ ᵱ}> or 
<{\small \fontspec{CharisSIL}ƚ ɨ ɉ ɍ}>, or also the 
example of <ø> given at the start of this section. Second, the same situation also 
occurs with combining characters designated as ``hook'', like < {\large  ̡}> 
\textsc{combining palatalized hook below} at \uni{0321}. This element seems to
occur in precomposed characters like <{\small \fontspec{CharisSIL}ᶀ ᶁ ᶂ ᶄ}>. 
However, all these characters are not canonical equivalent with a combination of 
their base characters with the combining overlay or combining hook.

@@

A central principle of Unicode is the distinction between character and glyph. A
character is the abstract notion of a symbol in a writing system, while a glyph
is the concrete drawing of such a symbol. In practice, there is a complex
interaction between characters and glyphs. A single Unicode character may of
course be rendered as a single glyph. However, a character may also be a
\textit{piece} of a glyph, or vice-versa. Actually, all possible relations
between glyphs and characters are attested.

First, \textit{a single character may have different contextually determined
glyphs.} For example, characters in writing systems like Hebrew and Arabic have
different glyphs depending on where they appear in a word. Some letters in
Hebrew change their form at the end of the word, and in Arabic, primary letters
have four contextually-sensitive variants (isolated, word initial, medial and
final). Second, \textit{a single character may be rendered as a sequence of
multiple glyphs.} For example, in Tamil one Unicode character may result in a
combination of a consonant and vowel, which are rendered as two adjacent glyphs
by fonts that supports Tamil. Third, \textit{a single glyph may be a combination
of multiple characters.} For example, the ligature <ï¬>, a single glyph, is the
result of two characters, <f> and <i>, that have undergone glyph substitution by
font rendering. Like contextually-determined glyphs, ligatures are (intended)
artifacts of text processing instructions. Finally, \textit{a single glyph may
be a part of a character} (e.g.~diacritics).

Further, the rendering of a glyph is also always dependent on the font being
used. For example, the Unicode character \textsc{latin small letter g} appears
as and in the Computer Modern and Courier fonts because their typefaces are
designed differently. Furthermore, font face may change the visual appearance of
a character, for example Times New Roman two-story <a> changes to a single-story
a in italics <\emph{a}>. This becomes a real problem for some phonetic
typesetting (see Section~\ref{pitfall-missing-glyphs}).

In sum, character-to-glyph mappings are complex technological issues that the
Unicode Consortium has had to address in the development of the Unicode
Standard, but for the lay user they can be utterly confusing because visual
rendering does not indicate logical encoding, or, in other words, what you see
isn't (necessarily) what you get.

>>>>

The other problem with visual display is related to the so-called \textsc{font
rendering}. Font rendering refers to the process of the actual positioning of
Unicode characters on a page of written text. This positioning is actually a
highly complex problem, and many things can go wrong in the process. Well-known
rendering problems, like proportional glyph size or ligatures are reasonably
well understood. However, even within the Latin script, the positioning of
diacritics relative to a base character is still a widespread problem.
Especially when more than one diacritic is supposed to be placed above (or
below) each other, this often leads to unexpected effects in many modern
software applications. The problems arising in Arabic and in many southeast
Asian scripts (like Devanagari or Burmese) are even much more complex. There are
basically three different approaches to font rendering. The most widespread is
Adobe's and Microsoft's \textsc{OpenType} system. This approach makes it
relatively easy for font developers, as the font itself does not include all
details about the precise placement of individual characters. For those details,
additional script-descriptions are necessary. All of those systems can lead to
unexpected behavior.\footnote{For more details about OpenType, see
\url{http://www.adobe.com/products/type/opentype.html} and
\url{http://www.microsoft.com/typography/otspec/}. Additional systems for
complex text layout are, among others, Microsoft's DirectWrite
\url{https://msdn.microsoft.com/library/dd368038.aspx} and the open-source
project HarfBuzz \url{http://www.freedesktop.org/wiki/Software/HarfBuzz/}.}
Alternative systems are Apple's \textsc{Apple Advanced Typography} (AAT) and the
open-source \textsc{Graphite} system from the Summer Institute of Linguistics
(SIL).\footnote{More information about AAT can be found at
\url{https://developer.apple.com/fonts/}. SIL's Graphite is described in detail
at
\url{http://scripts.sil.org/cms/scripts/page.php?site\_id=projects\&item\_id=graphite\_home}.}
In both of these systems, a larger burden is placed on the description inside
the font.

@@

\begin{itemize}
	\item[] <{\fontspec{EversonMono}A}> \textsc{latin capital letter a}, at \uni{0041} 
	\item[] <{\fontspec{EversonMono}А}> \textsc{cyrillic capital letter a}, at \uni{0410} 
	\item[] <{\fontspec{EversonMono}Α}> \textsc{greek capital letter alpha}, at \uni{0391} 
	\item[] <{\fontspec{EversonMono}Ꭺ}> \textsc{cherokee letter go}, at \uni{13AA} 
	\item[] <{\fontspec{EversonMono}ᗅ}> \textsc{canadian syllabics carrier gho}, at \uni{15C5} 
	\item[] <{\fontspec{EversonMono}ᴀ}> \textsc{latin small letter capital a}, at \uni{1D00} 
	\item[] <{\fontspec{EversonMono}ꓮ}> \textsc{lisu letter a}, at \uni{A4EE} 
%	\item[] <{\fontspec{EversonMono}Ａ}> \textsc{fullwidth latin capital letter a}, at \uni{FF21} 
	\item[] <{\fontspec{EversonMono}𐊠}> \textsc{carian letter a}, at \uni{102A0} 
	\item[] <{\fontspec{EversonMono}𐌀}> \textsc{old italic letter a}, at \uni{10300} 
	\item[] <{\fontspec{EversonMono}𝖠}> \textsc{mathematical sans-serif capital a}, \uni{1D5A0} 
	\item[] <{\fontspec{EversonMono}𝙰}> \textsc{mathematical monospace capital a}, at \uni{1D670} 
\end{itemize}

@@

Consider for example the characters and following Unicode code points:
\begin{itemize}
	\def\labelenumi{\arabic{enumi}.} 
	\item[] <Å> \textsc{latin capital letter a with ring above} \uni{00C5} 
	\item[] <Å> \textsc{angstrom sign} \uni{212B}
	\item[] <Å> \textsc{latin capital letter a} \uni{0041}
	+ \textsc{combining ring above} \uni{030A}
\end{itemize}

@@

There are two combinations that deserve specific care because it is actually 
possible to produce identical characters in different ways, without them being 
canonical equivalent. First, there are the
combining characters designated as ``overlay'' in the Unicode Standard, like    
<\ \ {\large  ̴}> \textsc{combining tilde overlay} at \uni{0334} or 
<\ \ {\large  ̵}> \textsc{combining short stroke overlay} at \uni{0335}. 
There are many characters that look like they are 
precomposed with such an overlay, for example 
<\charis{ɫ ᵬ ᵭ ᵱ}> or 
<\charis{ƚ ɨ ɉ ɍ}>, or also the 
example of <ø> given at the start of this section. Second, the same situation also 
occurs with combining characters designated as ``hook'', like < {\large  ̡}> 
\textsc{combining palatalized hook below} at \uni{0321}. This element seems to
occur in precomposed characters like <\charis{ᶀ ᶁ ᶂ ᶄ}>. 
However, all these characters are not canonical equivalent with a combination of 
their base characters with the combining overlay or combining hook.

@@

A central principle of Unicode is the distinction between character and glyph. A
character is the abstract notion of a symbol in a writing system, while a glyph
is the concrete drawing of such a symbol. In practice, there is a complex
interaction between characters and glyphs. A single Unicode character may of
course be rendered as a single glyph. However, a character may also be a piece
of a glyph, or vice-versa. Actually, all possible relations between glyphs and
characters are attested.

First, a single character may have different contextually determined glyphs. For
example, characters in writing systems like Hebrew and Arabic have different
glyphs depending on where they appear in a word. Some letters in Hebrew change
their form at the end of the word, and in Arabic, primary letters have four
contextually-sensitive variants (isolated, word initial, medial and final).
Second, a single character may be rendered as a sequence of multiple glyphs. For
example, in Tamil one Unicode character may result in a combination of a
consonant and vowel, which are rendered as two adjacent glyphs by fonts that
supports Tamil. Third, a single glyph may be a combination of multiple
characters. For example, the ligature <ﬁ>, a single glyph, is the result of two
characters, <f> and <i>, that have undergone glyph substitution by font
rendering (see also Section~\ref{pitfall-faulty-rendering}). Like
contextually-determined glyphs, ligatures are (intended) artifacts of text
processing instructions. Finally, a single glyph may be a part of a
character, as exemplified by diacritics.

Further, the rendering of a glyph is also always dependent on the font being
used. For example, the Unicode character \textsc{latin small letter g} appears
as <g> and <{\fontspec{Courier}g}> in the Linux Libertine and Courier fonts,
respectively, because their typefaces are designed differently. Furthermore,
font face may change the visual appearance of a character, for example Times New
Roman two-story <a> changes to a single-story a in italics <\emph{a}>. This
becomes a real problem for some phonetic typesetting (see
Section~\ref{pitfall-ipa-homoglyphs}).

In sum, character-to-glyph mappings are complex technological issues that the
Unicode Consortium has had to address in the development of the Unicode
Standard, but for the lay user they can be utterly confusing because visual
rendering does not (necessarily) indicate logical encoding.
